---
title: Mode
author: Brendan Knapp
date: '2018-11-23'
slug: index
categories:
  - data-science-from-scratch
  - statistics
  - central-tendency
  - tutorial
  - programming
tags:
  - data-science-from-scratch
  - statistics
  - programming
  - r
  - python
  - rcpp
image:
  caption: ''
  focal_point: 'Smart'
summary: 'Finding the mode of a sequence, from scratch.'
output:
  blogdown::html_page:
    toc: true
    toc_depth: 5
citation_url: https://knapply.com/post/mode/
editor_options: 
  chunk_output_type: console
---

> _This post was last updated on `r format(Sys.Date(), "%Y-%m-%d")`._

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, comment = "#>", message = FALSE, warning = FALSE)

options(width=90)

source(here::here("static/R/html-helpers.R"))
source(here::here("static/R/citation.R"))
source(here::here("static/R/wrangle_plot_save.R"))
```



# What is it?

Given a sequence, the mode is the value with the highest number of occurrences.

* Steps
  + find the unique values
  + count the number of occurrences of each value
  + the mode is the value with the highest number of occurrences

# R

## Base R

```{r}
r_mode <- function(x) {
  uniques <- x[!duplicated(x)]
  uniques[which.max(tabulate(match(x, uniques)))]
}


r_vector <- c(1, 4, 3, 2, 4, 5, 4, 3, 4, 6, 4, 3)

r_mode(r_vector)
```


# Rcpp

## Blueprint

```{Rcpp}
//[[Rcpp::plugins(cpp17)]]
#include <Rcpp.h>

// [[Rcpp::export]]
double cpp_mode_blueprint(const Rcpp::NumericVector& x) {
  std::unordered_map<double, std::size_t> counter;
  
  auto max_count = 0;
  double out = 0;
  for (auto i : x) {
    counter[i]++;
    if (counter[i] > max_count) {
      max_count = counter[i];
      out = i;
    }
  }
  
  return out;
}
```

```{r}
r_vector <- c(4, 1, 4, 3, 2, 5, 4, 3, 4, 6, 4, 3)

cpp_mode_blueprint(r_vector)
```



## Templated 

```{Rcpp}
//[[Rcpp::plugins(cpp17)]]
#include <Rcpp.h>

template <int RTYPE, typename T>
auto cpp_mode_(const Rcpp::Vector<RTYPE>& x) {
  std::unordered_map<T, size_t> counter(std::size(x));
  
  for (auto i : x) {
    counter[i]++;
  }
  
  size_t max = 0;
  T out;
  for (const auto& i : counter) {
    if (i.second > max) {
      max = i.second;
      out = i.first;
    }
  }

  return Rcpp::wrap(out);
}


// [[Rcpp::export]]
SEXP cpp_mode(const SEXP& x) {
  switch( TYPEOF(x) ) {
    case INTSXP:
      return cpp_mode_<INTSXP, int>(x);
    case REALSXP:
      return cpp_mode_<REALSXP, double>(x);
    case LGLSXP:
      return cpp_mode_<LGLSXP, bool>(x);
    case STRSXP:
      return cpp_mode_<STRSXP, Rcpp::String>(x);
  }
  Rcpp::warning("Invalid SEXPTYPE %d (%s).\n", TYPEOF(x), Rcpp::type2name(x));
  return R_NilValue;
}
```

```{r}
r_vector <- c(4, 1, 4, 3, 2, 5, 4, 3, 4, 6, 4, 3)

cpp_mode(r_vector)
```


# Python

## Pure Python

```{python}
import statistics

py_list = [4, 1, 4, 3, 2, 5, 4, 3, 4, 6, 4, 3]

res = statistics.mode(py_list)
print(res)
```


## NumPy

```{python}
import numpy as np

def numpy_mode(x):
    keys, counts = np.unique(x, return_counts = True)
    return keys[counts.argmax()]
    
numpy_array = np.array(py_list)

res = numpy_mode(numpy_array)
print(res)
```

## Numba

```{python}
import numba

@numba.jit()
def numba_mode(x):
  keys, counts = np.unique(x, return_counts = True)
  return keys[counts.argmax()]

res = numba_mode(numpy_array)
print(res)
```


### Parallel

```{python}
import numba

@numba.jit(parallel = True)
def numba_par_mode(x):
  keys, counts = np.unique(x, return_counts = True)
  return keys[counts.argmax()]

res = numba_par_mode(numpy_array)
print(res)
```

## CuPy


```{python}
import cupy as cp

def cupy_mode(x):
  keys, counts = cp.unique(x, return_counts = True)
  return keys[counts.argmax()]
  
cupy_array = cp.array(numpy_array, copy = True)

res = cupy_mode(cupy_array)
print(res)
```

# Shootout

<details>
<summary> Test Setup </summary>
<p>


```{r, message=FALSE, warning=FALSE}
library(reticulate)
library(tidyverse)
library(bench)
```

Create R Data

```{r}
set.seed(831)
r_data <- seq(-1e4, 1e4) %>%
  sample(replace = TRUE, size = 1e7) %>% 
  c(rep(32L, 200)) %>%  # our mode is 32
  sample()

validation_data <- r_data
```

Transfer to Python

```{python}
from copy import deepcopy

numpy_data = np.array(deepcopy(r.r_data))
cupy_data = cp.array(numpy_data, copy = True)
```


Do all the data match?

```{r}
are_equal <- list(r_data, py$numpy_data,
                  py$cp$asnumpy(py$cupy_data)) %>% 
    map_lgl(~ all.equal(validation_data, as.vector(.x)))

stopifnot(
  all(are_equal)
)
```

```{r}
(n <- scales::comma(length(r_data)))
(mem <- capture.output(pryr::object_size(r_data)))
```

</p>
</details>


<details>
<summary> `bench::mark()` </summary>
<p>

```{r, include=FALSE}
gc()
```


```{r}
res <- mark(
  r_mode(r_data),
  cpp_mode(r_data),
  py_run_string("numpy_mode(numpy_data)", convert = FALSE),
  py_run_string("numba_mode(numpy_data)", convert = FALSE),
  py_run_string("numba_par_mode(numpy_data)", convert = FALSE),
  py_run_string("cupy_mode(cupy_data)", convert = FALSE),
  min_iterations = 5,
  check = FALSE, filter_gc = FALSE
)
```

</p>
</details>


```{r, echo=FALSE}
as_chunk(here::here("static/R/wrangle_plot_save.R"))
```

<details>
<summary> `wrangle_plot_save()` </summary>
<p>

```{r wrangle_plot_save, eval=FALSE}
```

</p>
</details>

<details>
<summary> `write_gg()` </summary>
<p>

```{r write_gg, eval=FALSE}
```

</p>
</details>


<details>
<summary> `res_gg` </summary>
<p>

```{r, fig.align='center', fig.width=9, fig.height=11, message=FALSE, warning=FALSE}
res_gg <- wrangle_plot_save(res, .n = n, .mem = mem, 
                            .title = "Benchmarks: Statistical Mode")
```



```{r, eval=FALSE}
res_gg
```

</p>
</details>


```{r, echo=FALSE, fig.align='center', fig.width=9, fig.height=10, message=FALSE, warning=FALSE}
res_gg
```


<details>
<summary> Post Test Checks </summary>
<p>

Do all methods obtain the same results?

```{r}
control_res <- 32L

c(
  r_mode(r_data),
  cpp_mode(r_data),
  py$numpy_mode(py$numpy_data),
  py$numba_mode(py$numpy_data),
  py$numba_par_mode(py$numpy_data),
  py$cp$asnumpy(py$cupy_mode(py$cupy_data))
  )

are_equal <- c(
  
  r_mode(r_data),
  cpp_mode(r_data),
  py$numpy_mode(py$numpy_data),
  py$numba_mode(py$numpy_data),
  py$numba_par_mode(py$numpy_data),
  py$cp$asnumpy(py$cupy_mode(py$cupy_data))
  
  ) %>% 
  map_lgl(~ control_res == .x)

stopifnot(
  all(are_equal)
)
```


Did the input data remain unchanged?

```{r}
are_equal <- list(r_data, 
                  py$numpy_data,
                  py$cp$asnumpy(py$cupy_data)) %>% 
    map_lgl(~ all.equal(validation_data, as.vector(.x)))

stopifnot(
  all(are_equal)
)
```

</p>
</details>

```{r child=here::here("static/R/environment-details.Rmd")}
```

