---
title: Mode
author: Brendan Knapp
date: '2018-11-23'
slug: index
categories:
  - data-science-from-scratch
  - statistics
  - central-tendency
  - tutorial
  - programming
tags:
  - data-science-from-scratch
  - statistics
  - programming
  - r
  - python
  - rcpp
image:
  caption: ''
  focal_point: 'Smart'
summary: 'Finding the mode of a sequence, from scratch.'
output:
  blogdown::html_page:
    toc: true
    toc_depth: 5
citation_url: https://knapply.com/post/mode/
editor_options: 
  chunk_output_type: console
---

> _This post was last updated on `r format(Sys.Date(), "%Y-%m-%d")`._

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, comment = "#>", message = FALSE, warning = FALSE)

options(width=90)

source(here::here("static/R/html-helpers.R"))
source(here::here("static/R/citation.R"))
```




# What is it?

Given a sequence, the mode is the value with the highest number of occurrences.

* Steps
  + find the unique values
  + count the number of occurrences of each value
  + the mode is the value with the highest number of occurrences

# R


```{r}
r_mode <- function(x) {
  uniques <- x[!duplicated(x)]
  uniques[which.max(tabulate(match(x, uniques)))]
}


r_vector <- c(1, 4, 3, 2, 4, 5, 4, 3, 4, 6, 4, 3)

r_mode(r_vector)
```


# Rcpp

```{Rcpp, ref.label=knitr::all_rcpp_labels(), include=FALSE}
```

__Note:__ `Rcpp` examples assume code is prefaced with...

```{Rcpp, eval=FALSE}
//[[Rcpp::plugins(cpp17)]]
#include <Rcpp.h>
using namespace Rcpp;
```

> C++ References

* [`[]()`](https://en.cppreference.com/w/cpp/language/lambda) (Lambda)
* [`auto`](https://en.cppreference.com/w/cpp/language/auto)
* [`<algorithm>`](https://en.cppreference.com/w/cpp/header/algorithm)
    + [`std::max_element()`](https://en.cppreference.com/w/cpp/algorithm/max_element)
* [`<iterator>`](https://en.cppreference.com/w/cpp/header/iterator)
    + [`std::begin()`](https://en.cppreference.com/w/cpp/iterator/begin)
    + [`std::end()`](https://en.cppreference.com/w/cpp/iterator/end)
* [`<unordered_map>`](https://en.cppreference.com/w/cpp/header/unordered_map)
    + [`std::unordered_map`](https://en.cppreference.com/w/cpp/container/unordered_map)
* [`<utility>`](https://en.cppreference.com/w/cpp/header/utility)
    + [`std::pair`](https://en.cppreference.com/w/cpp/utility/pair)
    

## Blueprint

```{Rcpp, eval=FALSE}
// [[Rcpp::export]]
double cpp_mode_blueprint(const NumericVector& x) {
  std::unordered_map<double, size_t> counts;

  for (auto i : x) {
    counts[i]++;
  }
  
  auto compare_map_vals = [] (const std::pair<double, size_t>& p1, 
                              const std::pair<double, size_t>& p2) {
    return p1.second < p2.second;
  };

  const auto max_index = std::max_element(std::begin(counts), std::end(counts), 
                                          compare_map_vals);

  return (*max_index).first;
}
```

```{r}
r_vector <- c(4, 1, 4, 3, 2, 5, 4, 3, 4, 6, 4, 3)

cpp_mode_blueprint(r_vector)
```

## Template

```{Rcpp, eval = FALSE}
template <int RTYPE, typename T>
auto __cpp_mode(const Vector<RTYPE>& x) {
  std::unordered_map<T, size_t> counts;
  
  for (auto i : x) {
    counts[i]++;
  }
  
  auto compare_map_vals = [] (const std::pair<T, size_t>& p1,
                              const std::pair<T, size_t>& p2) {
    return p1.second < p2.second;
    };
  
  const auto max_index = std::max_element(std::begin(counts), std::end(counts), 
                                          compare_map_vals);
  
  return wrap( (*max_index).first );
}


// [[Rcpp::export]]
SEXP cpp_mode(const SEXP& x) {
  switch( TYPEOF(x) ) {
    case INTSXP:
      return __cpp_mode<INTSXP, int>(x);
    case REALSXP:
      return __cpp_mode<REALSXP, double>(x);
    case LGLSXP:
      return __cpp_mode<LGLSXP, bool>(x);
    case STRSXP:
      return __cpp_mode<STRSXP, String>( x );
    default: 
      stop("Invalid SEXPTYPE %d (%s).\n", TYPEOF(x), type2name(x));
  }
  return R_NilValue;
}
```

```{r}
r_vector <- c(4, 1, 4, 3, 2, 5, 4, 3, 4, 6, 4, 3)

cpp_mode(r_vector)
```


# Python

## `statistics.mode()` (Built-in)

```{python}
import statistics

py_list = [4, 1, 4, 3, 2, 5, 4, 3, 4, 6, 4, 3]

res = statistics.mode(py_list)
print(res)
```

## Custom Pure Python

```{python}
from collections import Counter

def py_mode(x):
  return Counter(x).most_common(1)[0][0]

res = py_mode(py_list)
print(res)
```

## Custom NumPy

```{python}
import numpy as np

def numpy_mode(x):
    keys, counts = np.unique(x, return_counts = True)
    return keys[counts.argmax()]
    
numpy_array = np.array(py_list)

res = numpy_mode(numpy_array)
print(res)
```



# Shootout

<details>
<summary> Test Setup </summary>
<p>


```{r, message=FALSE, warning=FALSE}
library(reticulate)
library(tidyverse)
library(bench)
```

Create R Data

```{r}
resample <- function(x, .n = 30) {
  `dim<-`(replicate(n = .n, sample(x, replace = TRUE)), NULL)
}

set.seed(831)
big_dbl <- seq(-1e7, 1e7, by = 0.02) %>%
  sample(replace = TRUE, size = 1e5) %>%
  c(rep(32.4, 100)) %>%  # our mode is 32.4
  resample()

validation_data <- big_dbl
```

Transfer to Python

```{python}
py_big_dbl = r.big_dbl
numpy_big_dbl = np.array(py_big_dbl)
```


Do all the data match?

```{r}
stopifnot(
  all.equal(validation_data,
            big_dbl,
            py$py_big_dbl,
            py$numpy_big_dbl)
  )
```

```{r, eval=FALSE}
cat("sample:\t", paste0(head(big_dbl, 3), collapse = ", "), " ... ",
    paste0(tail(big_dbl, 3), collapse = ", "),  "\n",
    "n = ", scales::comma(length(big_dbl)), "\n",
    "memory footprint = ", capture.output(pryr::object_size(big_dbl)), sep = "")
```

</p>
</details>

```{r, echo=FALSE}
cat("sample:\t", paste0(head(big_dbl, 3), collapse = ", "), " ... ",
    paste0(tail(big_dbl, 3), collapse = ", "),  "\n",
    "n = ", scales::comma(length(big_dbl)), "\n",
    "memory footprint = ", capture.output(pryr::object_size(big_dbl)), sep = "")
```


<details>
<summary> Run Benchmark Test </summary>
<p>

```{r, echo=FALSE}
gc()
```


```{r}
res <- mark(
  r_mode(big_dbl),
  cpp_mode(big_dbl),
  py_run_string("py_mode(py_big_dbl)", convert = FALSE),
  py_run_string("statistics.mode(py_big_dbl)", convert = FALSE),
  py_run_string("numpy_mode(numpy_big_dbl)", convert = FALSE),
  iterations = 50, check = FALSE, filter_gc = FALSE
)
```


</p>
</details>






## Results

<details>
<summary> Plot Results </summary>
<p>

```{r, fig.align='center', fig.width=9, fig.height=10, message=FALSE, warning=FALSE}
res_gg <- res %>%
  mutate(expression = as.character(expression)) %>%
  mutate(type = recode(
    expression,
    "r_mode(big_dbl)" = "Custom R",
    "cpp_mode(big_dbl)" = "Custom Rcpp",
    'py_run_string("statistics.mode(py_big_dbl)", convert = FALSE)' = "Python's built-in",
    'py_run_string("numpy_mode(numpy_big_dbl)", convert = FALSE)' = "Custom NumPy",
    'py_run_string("py_mode(py_big_dbl)", convert = FALSE)' = "Custom Python"
  )) %>%
  mutate(lang = if_else(str_detect(type, "\\sR(cpp)?"), "R", "Python")) %>% 
  mutate(expression = recode(
    expression,
    'py_run_string("statistics.mode(py_big_dbl)", convert = FALSE)' = 'statistics.mode(py_big_dbl)',
    'py_run_string("numpy_mode(numpy_big_dbl)", convert = FALSE)' = 'numpy_mode(numpy_big_dbl)',
    'py_run_string("py_mode(py_big_dbl)", convert = FALSE)' = 'py_mode(py_big_dbl)'
  )) %>%
  unnest() %>%
  group_by(type) %>%
  mutate(med = median(as.numeric(time))) %>%
  ungroup() %>%
  arrange(desc(med)) %>%
  mutate(type = as_factor(type)) %>%
  arrange(med) %>%
  mutate(expression = as_factor(expression)) %>%
  
  ggplot(aes(x = type, y = time, color = expression)) +
    ggbeeswarm::geom_beeswarm(cex = 0.65) +
    guides(color = guide_legend(ncol = 2, override.aes = list(size = 3))) +
    coord_flip() +
    theme_minimal(base_size = 16, base_family = "serif") +
    theme(axis.title = element_blank(),
          legend.position = "bottom", legend.title = element_blank(),
          legend.text = element_text(family = "mono"),
          plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(face = "italic")) +
    labs(title = "Speed Shootout", subtitle = "Statistical Mode", caption = "knapply.com")

res_gg + facet_wrap(~ lang, ncol = 1, scales = "free_y")
```


```{r, eval=FALSE}
cat("sample:\t", paste0(head(big_dbl, 3), collapse = ", "), " ... ",
    paste0(tail(big_dbl, 3), collapse = ", "),  "\n",
    "n = ", scales::comma(length(big_dbl)), "\n",
    "memory footprint = ", capture.output(pryr::object_size(big_dbl)), sep = "")

res_gg
```

</p>
</details>


```{r, echo=FALSE, fig.align='center', fig.width=9, fig.height=10, message=FALSE, warning=FALSE}
res_gg
```



<details>
<summary> Post Test Checks </summary>
<p>

Do all methods obtain the same results?

```{r}
stopifnot(
  all.equal(r_mode(big_dbl), 
            cpp_mode(big_dbl), 
            py$numpy_mode(py$numpy_big_dbl),
            py$statistics$mode(py$py_big_dbl))
  )
```

Did the input data remain unchanged?

```{r}
stopifnot(
  all.equal(validation_data, 
            big_dbl, 
            py$py_big_dbl,
            py$numpy_big_dbl)
  )
```

</p>
</details>

## Environment

<details>
<summary> R </summary>
<p>

```{r}
sessionInfo()
```

</p>
</details>


<details>
<summary> C++ </summary>
<p>

```{r}
if (R.version["minor"] == "6.0") {
  "GCC 8.2.0"
  } else {
  system("gcc --version", intern = TRUE)[[1]]
  }
```

</p>
</details>


<details>
<summary> Python </summary>
<p>

```{python}
import sys

print(sys.version)
print(np.__version__)
```

</p>
</details>

<details>
<summary> CPU </summary>
<p>


```{r}
cat(system("wmic cpu get name", intern = TRUE)[[2]])
```

</p>
</details>


<details>
<summary> Memory </summary>
<p>

```{r}
system("wmic MEMORYCHIP get BankLabel, Capacity, Speed", intern = TRUE) %>%
  str_trim() %>%
  as_tibble() %>%
  slice(2:3) %>%
  separate(value, into = c("BankLabel", "Capacity", "Speed"), sep = "\\s{2,}") %>%
  mutate_at(vars(Capacity), ~ paste(round(as.numeric(.) / 1e9, 2), "GB")) %>%
  mutate(Speed = paste(Speed, "MHz")) %>%
  rename_all(str_replace, "L", " L") %>%
  knitr::kable(format = "html") %>%
  kableExtra::kable_styling(full_width = FALSE)
```

</p>
</details>


<br><br><br>

```{r, echo=FALSE, comment=NA, results='markup'}
print_citation_header()
```

```{r, echo=FALSE, comment=NA, results='markup', class.output='citation'}
print_citation()
print_bibtex()
```



