---
title: Median
author: Brendan Knapp
date: '2018-11-12'
slug: index
categories:
  - data-science-notes
  - statistics
  - central-tendency
tags:
  - data-science-notes
  - statistics
  - central-tendency
  - r
  - python
  - rcpp
image:
  caption: ''
  focal_point: 'Smart'
summary: 'Finding median of a sequence.'
output:
  blogdown::html_page:
    toc: true
    toc_depth: 5
citation_url: https://knapply.com/post/median/
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, comment = "#>", message = FALSE, warning = FALSE)

options(width=90)

source(here::here("static/R/html-helpers.R"))
source(here::here("static/R/citation.R"))
```



# What is it?

Given a sequence, which is the middle-most value. Put another way, what value is equally likely to be greater than the first half of a sorted sequence and lesser than the second half of a sorted sequence.

* Steps
  + sort the sequence
  + if the length of the sequence is odd
      + the middle value is the median
  + if length of the sequence is even
      + the mean of the 2 middle values is the median


# R

## `stats::median()`

```{r}
r_even_n <- c(126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
              190, 41, 612, 826, 507, 105, 14, 237, 669, 7)

r_odd_n <- c(126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
             190, 41, 612, 826, 507, 105, 14, 237, 669, 7, 375)

median(r_even_n)
median(r_odd_n)
```

## Custom R

_Note: This is essentially a stripped-down version of `median.default()`._

```{r}
r_median <- function(x) {
  n <- length(x)
  mid <- (n + 1L) %/% 2L
  partially_sorted <- sort(x, partial = n %/% 2L + 1L)
  if (n %% 2L) {
    return(partially_sorted[mid])
  }      
  ( partially_sorted[[mid]] + partially_sorted[[mid + 1L]] ) / 2L 
}

r_median(r_even_n)
r_median(r_odd_n)
```


# Rcpp

```{Rcpp, ref.label=knitr::all_rcpp_labels(), include=FALSE}
```

__Note:__ `Rcpp` examples assume code is prefaced with...

```{Rcpp, eval=FALSE}
//[[Rcpp::plugins(cpp17)]]
#include <Rcpp.h>
using namespace Rcpp;
```


## `Rcpp::median()`

```{Rcpp, eval=FALSE}
// [[Rcpp::export]]
double rcpp_sugar_median(NumericVector x) {
  return median(x);
}
```

```{r}
r_even_n <- c(126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
              190, 41, 612, 826, 507, 105, 14, 237, 669, 7)

r_odd_n <- c(126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
             190, 41, 612, 826, 507, 105, 14, 237, 669, 7, 375)

rcpp_sugar_median(r_even_n)
rcpp_sugar_median(r_odd_n)
```

## Custom

### Blueprint

```{Rcpp, eval=FALSE}
using std::begin;
using std::end;
using std::nth_element;
using std::size;

// [[Rcpp::export]]
double cpp_median_blueprint(const NumericVector& x) {
  auto temp = clone(x);
  const auto n = size(x);
  auto mid = begin(temp) + n / 2;
  
  std::nth_element( begin(temp), mid, end(temp) );

  if (n % 2) {
    return *mid;
  }

  const auto mid1 = *mid;
  nth_element( begin(temp), --mid, end(temp) );

  return ( mid1 + *mid ) / 2.0;
}
```




```{r}
cpp_median_blueprint(r_even_n)
cpp_median_blueprint(r_odd_n)
```

### Template Version

```{Rcpp, eval=FALSE}
template <int RTYPE>
auto cpp_median_(const Vector<RTYPE>& x) {
  auto temp = clone(x);
  const auto n = std::size(temp);
  auto mid = std::begin(temp) + n / 2;

  std::nth_element( std::begin(temp), mid, std::end(temp) );

  if (n % 2) {
    return wrap( *mid );
  }
  
  const auto mid1 = *mid;
  std::nth_element( std::begin(temp), --mid, std::end(temp) );

  return wrap( (mid1 + *mid) / 2.0 );
}

// [[Rcpp::export]]
SEXP cpp_median(const SEXP& x) {
  switch( TYPEOF(x) ) {
    case INTSXP:
      return cpp_median_<INTSXP>(x);
    case REALSXP:
      return cpp_median_<REALSXP>(x);
  }
  warning("Invalid SEXPTYPE %d (%s).\n", TYPEOF(x), type2name(x));
  return R_NilValue;
}
```


```{r}
cpp_median(r_even_n)
cpp_median(r_odd_n)
```


# Python

## `statistics.median()`

Python's `statistics` module is part of its standard library.

```{python}
import statistics

py_even_n = [126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
             190, 41, 612, 826, 507, 105, 14, 237, 669, 7]

py_odd_n = [126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
            190, 41, 612, 826, 507, 105, 14, 237, 669, 7, 375]

res1 = statistics.median(py_even_n)
res2 = statistics.median(py_odd_n)

print([res1, res2])
```

## `numpy.median()`

```{python}
import numpy as np

numpy_even_n = np.array(py_even_n)
numpy_odd_n = np.array(py_odd_n)

res1 = np.median(numpy_even_n)
res2 = np.median(numpy_odd_n)

print(res1, res2)
```

## Numba

```{python}
import numba

@numba.jit()
def numba_median(x):
  return np.median(x)

res1 = numba_median(numpy_even_n)
res2 = numba_median(numpy_odd_n)

print(res1, res2)
```

### Parallel

```{python}
@numba.jit(parallel=True)
def numba_parallel_median(x):
  return np.median(x)

res1 = numba_parallel_median(numpy_even_n)
res2 = numba_parallel_median(numpy_odd_n)

print(res1, res2)
```

## Custom CuPy

```{python}
import cupy as cp

def cupy_median(x):
  n = cp.size(x)
  sorted_x = cp.sort(x)
  if n % 2:
    return sorted_x[n // 2]
  return (sorted_x[n // 2 - 1] + sorted_x[n // 2]) / 2
  
cupy_even_n = cp.array(numpy_even_n)
cupy_odd_n = cp.array(numpy_odd_n)

res1 = cupy_median(cupy_even_n)
res2 = cupy_median(cupy_odd_n)

print(res1, res2)
```


## Custom Pure Python

```{python}
def py_median(x):
  n = len(x)
  sorted_x = sorted(x)
  if n % 2:
    return sorted_x[n // 2]
  return (sorted_x[n // 2 - 1] + sorted_x[n // 2]) / 2

res1 = py_median(py_even_n)
res2 = py_median(py_odd_n)

print([res1, res2])
```




# Shootout

<details>
<summary> Test Setup </summary>
<p>

```{r, message=FALSE, warning=FALSE}
library(reticulate)
library(tidyverse)
library(bench)
```

Create R Data

```{r}
set.seed(831)
big_dbl <- seq(-1e7, 1e7, by = 0.02) %>%
  sample(replace = TRUE, size = 1e6) %>%
  rep(30) %>%
  c(rep(32.4, 20)) %>% 
  sample(replace = TRUE)

validation_data <- big_dbl
```

Transfer to Python

```{python}
from copy import deepcopy

py_big_dbl = deepcopy(r.big_dbl)
numpy_big_dbl = np.array(deepcopy(py_big_dbl))
cupy_big_dbl = cp.array(deepcopy(numpy_big_dbl))
```


Do all the data match?

```{r}
stopifnot(
  all.equal(validation_data,
            big_dbl,
            py$py_big_dbl,
            py$numpy_big_dbl,
            cp$cupy_big_dbl)
  )
```

```{r}
(n <- scales::comma(length(big_dbl)))
(mem <- capture.output(pryr::object_size(big_dbl)))
```

</p>
</details>


<details>
<summary> Run Benchmark Test </summary>
<p>

```{r, include=FALSE}
gc()
```


```{r}
library(bench)

res <- mark(
  median.default(big_dbl),
  r_median(big_dbl),
  rcpp_sugar_median(big_dbl),
  cpp_median(big_dbl),
  py_run_string("statistics.median(py_big_dbl)", convert = FALSE),
  py_run_string("np.median(numpy_big_dbl)", convert = FALSE),
  py_run_string("py_median(numpy_big_dbl)", convert = FALSE),
  py_run_string("numba_median(numpy_big_dbl)", convert = FALSE),
  py_run_string("numba_parallel_median(numpy_big_dbl)", convert = FALSE),
  py_run_string("cupy_median(cupy_big_dbl)", convert = FALSE),
  min_iterations = 5,
  check = FALSE, filter_gc = FALSE
)
```

</p>
</details>


## Results

<details>
<summary> Plot </summary>
<p>

```{r, fig.align='center', fig.width=9, fig.height=11, message=FALSE, warning=FALSE}
res_gg <- res %>% 
  mutate(expression = as.character(expression)) %>% 
  mutate(type = recode(
    expression,
    "median.default(big_dbl)" = "R's built-in",
    "r_median(big_dbl)" = "Custom R",
    "rcpp_sugar_median(big_dbl)" = "Rcpp's built-in",
    "cpp_median(big_dbl)" = "Custom Rcpp",
    'py_run_string("statistics.median(py_big_dbl)", convert = FALSE)' = "Python's built-in",
    'py_run_string("np.median(numpy_big_dbl)", convert = FALSE)' = "NumPy",
    'py_run_string("py_median(numpy_big_dbl)", convert = FALSE)' = "Custom Python",
    'py_run_string("numba_median(numpy_big_dbl)", convert = FALSE)' = "Numba",
    'py_run_string("numba_parallel_median(numpy_big_dbl)", convert = FALSE)' = "Numba (Parallel)",
    'py_run_string("cupy_median(cupy_big_dbl)", convert = FALSE)' = "Custom CuPy (GPU)"
  )) %>%  
  mutate(expression = recode(
    expression,
    'py_run_string("statistics.median(py_big_dbl)", convert = FALSE)' = 'statistics.median(py_big_dbl)',
    'py_run_string("np.median(numpy_big_dbl)", convert = FALSE)' = 'np.median(numpy_big_dbl)',
    'py_run_string("py_median(numpy_big_dbl)", convert = FALSE)' = 'py_median(numpy_big_dbl)',
    'py_run_string("numba_median(numpy_big_dbl)", convert = FALSE)' = "numba_median(numpy_big_dbl)",
    'py_run_string("numba_parallel_median(numpy_big_dbl)", convert = FALSE)' = 'numba_parallel_median(numpy_big_dbl)',
    'py_run_string("cupy_median(cupy_big_dbl)", convert = FALSE)' = "cupy_median(cupy_big_dbl)"
  )) %>% 
  unnest() %>%
  group_by(type) %>%
  mutate(med = median(as.numeric(time))) %>%
  ungroup() %>%
  arrange(desc(med)) %>%
  mutate(type = as_factor(type)) %>%
  arrange(med) %>%
  mutate(expression = as_factor(expression)) %>%
  ggplot(aes(x = type, y = time, color = expression)) +
    # ggbeeswarm::geom_beeswarm(cex = 0.65) +
    stat_ydensity(aes(fill = expression), scale = "width", bw = 0.01, 
                  trim = FALSE) +
    guides(fill = guide_legend(ncol = 2, override.aes = list(size = 3))) +
    coord_flip() +
    theme_minimal(base_size = 16, base_family = "serif") +
    theme(axis.title = element_blank(),
          legend.position = "bottom", legend.title = element_blank(),
          legend.text = element_text(family = "mono"),
          plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(face = "italic"),
          plot.caption = element_text(family = "mono", size = 16, face = "bold")) +
    labs(title = "Speed Shootout: Median", 
         subtitle = glue::glue(
           "n = {n} ({mem})"
           ), 
         caption = "knapply.com")
```




```{r, eval=FALSE}
res_gg
```

</p>
</details>


```{r, echo=FALSE, fig.align='center', fig.width=9, fig.height=10, message=FALSE, warning=FALSE}
res_gg
```


<details>
<summary> Post Test Checks </summary>
<p>

Do all methods obtain the same results?

```{r, eval=FALSE}
stopifnot(
  all.equal(
    median.default(big_dbl),
    r_median(big_dbl),
    rcpp_sugar_median(big_dbl),
    cpp_median(big_dbl),
    py$np$median(py$numpy_big_dbl),
    py$statistics$median(py$py_big_dbl),
    py$py_median(py$py_big_dbl),
    py$numba_median(py$numpy_big_dbl),
    py$numba_parallel_median(py$numpy_big_dbl),
    py$cupy_median(py$cupy_big_dbl)
  )
)
```



Did the input data remain unchanged?

```{r}
stopifnot(
  all.equal(validation_data, 
            big_dbl, 
            py$py_big_dbl,
            py$numpy_big_dbl)
  )
```

</p>
</details>

## Environment

<details>
<summary> R </summary>
<p>

```{r}
sessionInfo()
```

</p>
</details>


<details>
<summary> C++ </summary>
<p>

```{r}
if (R.version["minor"] == "6.0") {
  "GCC 8.2.0"
  } else {
  system("gcc --version", intern = TRUE)[[1]]
  }
```

</p>
</details>


<details>
<summary> Python </summary>
<p>

```{python}
import sys

print(sys.version)
print(np.__version__)
print(numba.__version__)
print(cp.__version__)
```

</p>
</details>

<details>
<summary> CPU </summary>
<p>


```{r}
cat(system("wmic cpu get name", intern = TRUE)[[2]])
```

</p>
</details>

<details>
<summary> GPU </summary>
<p>

```{r}
system("wmic path win32_VideoController get name", intern = TRUE) %>% 
  str_subset("NVIDIA") %>% 
  str_trim()
```

</p>
</details>


<details>
<summary> Memory </summary>
<p>

```{r}
system("wmic MEMORYCHIP get BankLabel, Capacity, Speed", intern = TRUE) %>%
  str_trim() %>%
  as_tibble() %>%
  slice(2:3) %>%
  separate(value, into = c("BankLabel", "Capacity", "Speed"), sep = "\\s{2,}") %>%
  mutate_at(vars(Capacity), ~ paste(round(as.numeric(.) / 1e9, 2), "GB")) %>%
  mutate(Speed = paste(Speed, "MHz")) %>%
  rename_all(str_replace, "L", " L") %>%
  knitr::kable(format = "html") %>%
  kableExtra::kable_styling(full_width = FALSE)
```

</p>
</details>


<br><br><br>

```{r, echo=FALSE, comment=NA, results='markup'}
print_citation_header()
```

```{r, echo=FALSE, comment=NA, results='markup', class.output='citation'}
print_citation()
print_bibtex()
```

```{r, include=FALSE}

```

