---
title: Median
author: Brendan Knapp
date: '2018-11-12'
slug: index
categories:
  - data-science-notes
  - statistics
  - central-tendency
tags:
  - data-science-notes
  - statistics
  - central-tendency
  - r
  - python
  - rcpp
image:
  caption: ''
  focal_point: 'TopLeft'
summary: 'Finding median of a sequence.'
output:
  blogdown::html_page:
    toc: true
    toc_depth: 5
citation_url: https://knapply.com/post/median/
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, comment = "#>", message = FALSE, warning = FALSE)

options(width=90)

source(here::here("static/R/html-helpers.R"))
source(here::here("static/R/citation.R"))
source(here::here("static/R/wrangle_plot_save.R"))
```





# What is it?

Given a sequence, which is the middle-most value. Put another way, what value is equally likely to be greater than the first half of a sorted sequence and lesser than the second half of a sorted sequence.

* Steps
  + sort the sequence
  + if the length of the sequence is odd
      + the middle value is the median
  + if length of the sequence is even
      + the mean of the 2 middle values is the median


# R

## Base R

```{r}
r_even_n <- c(126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
              190, 41, 612, 826, 507, 105, 14, 237, 669, 7)

r_odd_n <- c(126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
             190, 41, 612, 826, 507, 105, 14, 237, 669, 7, 375)

median(r_even_n)
median(r_odd_n)
```

## Custom Base R

_Note: This is essentially a stripped-down version of `median.default()`._

```{r}
r_median <- function(x) {
  n <- length(x)
  mid <- (n + 1L) %/% 2L
  partially_sorted <- sort(x, partial = n %/% 2L + 1L)
  if (n %% 2L) {
    return(partially_sorted[mid])
  }      
  ( partially_sorted[[mid]] + partially_sorted[[mid + 1L]] ) / 2L 
}

r_median(r_even_n)
r_median(r_odd_n)
```


# Rcpp

## Sugar

```{Rcpp}
#include <Rcpp.h>

// [[Rcpp::export]]
double rcpp_sugar_median(Rcpp::NumericVector x) {
  return Rcpp::median(x);
}
```

```{r}
r_even_n <- c(126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
              190, 41, 612, 826, 507, 105, 14, 237, 669, 7)

r_odd_n <- c(126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
             190, 41, 612, 826, 507, 105, 14, 237, 669, 7, 375)

rcpp_sugar_median(r_even_n)
rcpp_sugar_median(r_odd_n)
```

## Custom

### Blueprint

```{Rcpp}
//[[Rcpp::plugins(cpp17)]]
#include <Rcpp.h>

// [[Rcpp::export]]
auto cpp_median_blueprint(const Rcpp::NumericVector& x) {
  auto temp = Rcpp::clone(x);
  const auto n = std::size(x);
  auto mid = std::begin(temp) + n / 2;
  
  std::nth_element( std::begin(temp), mid, std::end(temp) );

  if (n % 2) {
    return *mid;
  }

  const auto mid1 = *mid;
  std::nth_element( std::begin(temp), --mid, std::end(temp) );

  return ( mid1 + *mid ) / 2.0;
}
```




```{r}
cpp_median_blueprint(r_even_n)
cpp_median_blueprint(r_odd_n)
```

### Templated

```{Rcpp}
//[[Rcpp::plugins(cpp17)]]
#include <Rcpp.h>

template <int RTYPE>
auto cpp_median_(const Rcpp::Vector<RTYPE>& x) {
  auto temp = Rcpp::clone(x);
  const auto n = std::size(temp);
  auto mid = std::begin(temp) + n / 2;

  std::nth_element( std::begin(temp), mid, std::end(temp) );

  if (n % 2) {
    return Rcpp::wrap( *mid );
  }
  
  const auto mid1 = *mid;
  std::nth_element( std::begin(temp), --mid, std::end(temp) );

  return Rcpp::wrap( (mid1 + *mid) / 2.0 );
}

// [[Rcpp::export]]
SEXP cpp_median(const SEXP& x) {
  switch( TYPEOF(x) ) {
    case INTSXP:
      return cpp_median_<INTSXP>(x);
    case REALSXP:
      return cpp_median_<REALSXP>(x);
  }
  Rcpp::warning("Invalid SEXPTYPE %d (%s).\n", TYPEOF(x), Rcpp::type2name(x));
  return R_NilValue;
}
```


```{r}
cpp_median(r_even_n)
cpp_median(r_odd_n)
```


# Python

## Pure Python

```{python}
import statistics

py_even_n = [126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
             190, 41, 612, 826, 507, 105, 14, 237, 669, 7]

py_odd_n = [126, 793, 999, 525, 851, 798, 120, 714, 852, 151,
            190, 41, 612, 826, 507, 105, 14, 237, 669, 7, 375]

res1 = statistics.median(py_even_n)
res2 = statistics.median(py_odd_n)

print([res1, res2])
```

## NumPy

```{python}
import numpy as np

numpy_even_n = np.array(py_even_n)
numpy_odd_n = np.array(py_odd_n)

res1 = np.median(numpy_even_n)
res2 = np.median(numpy_odd_n)

print(res1, res2)
```

## Numba

```{python}
import numba

@numba.jit()
def numba_median(x):
  return np.median(x)

res1 = numba_median(numpy_even_n)
res2 = numba_median(numpy_odd_n)

print(res1, res2)
```

### Parallel

```{python}
@numba.jit(parallel=True)
def numba_par_median(x):
  return np.median(x)

res1 = numba_par_median(numpy_even_n)
res2 = numba_par_median(numpy_odd_n)

print(res1, res2)
```

## CuPy

```{python}
import cupy as cp

def cupy_median(x):
  n = cp.size(x)
  sorted_x = cp.sort(x)
  if n % 2:
    return sorted_x[n // 2]
  return (sorted_x[n // 2 - 1] + sorted_x[n // 2]) / 2
  
cupy_even_n = cp.array(numpy_even_n)
cupy_odd_n = cp.array(numpy_odd_n)

res1 = cupy_median(cupy_even_n)
res2 = cupy_median(cupy_odd_n)

print(res1, res2)
```




# Shootout

<details>
<summary> Test Setup </summary>
<p>


```{r, message=FALSE, warning=FALSE}
library(reticulate)
library(tidyverse)
library(bench)
```

Create R Data

```{r}
set.seed(831)
r_data <- seq(-1e4, 1e4, by = 0.2) %>%
  sample(replace = TRUE, size = 1e7) %>%
  sample()

validation_data <- r_data
```

Transfer to Python

```{python}
from copy import deepcopy

numpy_data = np.array(deepcopy(r.r_data))
cupy_data = cp.array(numpy_data, copy = True)
```


Do all the data match?

```{r}
are_equal <- list(r_data, py$numpy_data,
                  py$cp$asnumpy(py$cupy_data)) %>% 
    map_lgl(~ all.equal(validation_data, as.vector(.x)))

stopifnot(
  all(are_equal)
)
```

```{r}
(n <- scales::comma(length(r_data)))
(mem <- capture.output(pryr::object_size(r_data)))
```

</p>
</details>


<details>
<summary> `bench::mark()` </summary>
<p>

```{r, include=FALSE}
gc()
```


```{r}
res <- mark(
  median(r_data),
  rcpp_sugar_median(r_data),
  cpp_median(r_data),
  py_run_string("np.median(numpy_data)", convert = FALSE),
  py_run_string("numba_median(numpy_data)", convert = FALSE),
  py_run_string("numba_par_median(numpy_data)", convert = FALSE),
  py_run_string("cupy_median(cupy_data)", convert = FALSE),
  min_iterations = 5,
  check = FALSE, filter_gc = FALSE
)
```

</p>
</details>


```{r, echo=FALSE}
as_chunk(here::here("static/R/wrangle_plot_save.R"))
```

<details>
<summary> `wrangle_plot_save()` </summary>
<p>

```{r wrangle_plot_save, eval=FALSE}
```

</p>
</details>

<details>
<summary> `write_gg()` </summary>
<p>

```{r write_gg, eval=FALSE}
```

</p>
</details>


<details>
<summary> `res_gg` </summary>
<p>

```{r, fig.align='center', fig.width=9, fig.height=11, message=FALSE, warning=FALSE}
res_gg <- wrangle_plot_save(res, .n = n, .mem = mem, 
                            .title = "Benchmarks: Median")
```



```{r, eval=FALSE}
res_gg
```

</p>
</details>


```{r, echo=FALSE, fig.align='center', fig.width=9, fig.height=10, message=FALSE, warning=FALSE}
res_gg
```


<details>
<summary> Post Test Checks </summary>
<p>

Do all methods obtain the same results?

```{r}
(control_res <- median(r_data))

c(rcpp_sugar_median(r_data),
  cpp_median(r_data),
  py$np$median(py$numpy_data),
  py$numba_median(py$numpy_data),
  py$numba_par_median(py$numpy_data),
  py$cp$asnumpy(py$cupy_median(py$cupy_data)))

are_equal <- c(
  
  rcpp_sugar_median(r_data),
  cpp_median(r_data),
  py$np$median(py$numpy_data),
  py$numba_median(py$numpy_data),
  py$numba_par_median(py$numpy_data),
  py$cp$asnumpy(py$cupy_median(py$cupy_data))
  
  ) %>% 
  map_lgl(~ control_res == .x)

stopifnot(
  all(are_equal)
)
```


Did the input data remain unchanged?

```{r}
are_equal <- list(r_data, 
                  py$numpy_data,
                  py$cp$asnumpy(py$cupy_data)) %>% 
    map_lgl(~ all.equal(validation_data, as.vector(.x)))

stopifnot(
  all(are_equal)
)
```

</p>
</details>

```{r child=here::here("static/R/environment-details.Rmd")}
```

