---
title: "$|A|$: Determinant of a Matrix"
author: Brendan Knapp
date: '2018-12-11'
slug: index
categories:
  - data-science-notes
tags:
  - data-science-notes
  - linear-algebra
  - r
  - python
  - rcpp
image:
  caption: ''
  focal_point: 'Top'
summary: 'Finding the determinant of a matrix.'
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
citation_url: https://knapply.com/post/matrix-determinant/
editor_options: 
  chunk_output_type: console
---

> _This post was last updated on `r format(Sys.Date(), "%Y-%m-%d")`._

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, comment = "#>", message = FALSE, warning = FALSE)

options(width=90)

source(here::here("static/R/html-helpers.R"))
# source(here::here("static/R/citation.R"))
source(here::here("static/R/wrangle_plot_save.R"))
```


# Breakdown

## $2 \times 2$ Matrix

$$
A = \begin{bmatrix}
  a & b \\
  c & d
\end{bmatrix}
$$

$$
\left| \begin{matrix}
  \color{red}a & & &  & \color{blue}b \\
  &  \color{red}\nwarrow & &  \color{blue}\nearrow \\
  & & \large\times & &  \\
  &  \color{blue}\swarrow & &  \color{red}\searrow \\
  \color{blue}c & & & & \color{red}d \\
\end{matrix} \right|
$$

$$
|A| = \color{red}{ad} - \color{blue}{bc}
$$

## $3 \times 3$ Matrix

$$
A = \begin{bmatrix} a   &  b   &  c  \\
                    d   &  e   &  f  \\
                    g   &  h   &  i  \\ \end{bmatrix}\\
$$

Break it into three 2 x 2 matrices...

$$
\begin{bmatrix}
  \color{green}{\Large{\textbf{a}}} & \scriptsize{\color{gray}b} & \scriptsize{\color{gray}c} \\
  \scriptsize{\color{gray}d} & \color{red}e & \color{blue}f \\
  \scriptsize{\color{gray}g} & \color{blue}h & \color{red}i
\end{bmatrix}
\LARGE{-}
\normalsize
\begin{bmatrix}
  \scriptsize{\color{gray}a} & \color{green}{\Large{\textbf{b}}} & \scriptsize{\color{gray}c} \\
  \color{red}d & \scriptsize{\color{gray}e} & \color{blue}f \\
  \color{blue}g & \scriptsize{\color{gray}h} & \color{red}i
\end{bmatrix}
\LARGE+
\normalsize
\begin{bmatrix}
  \scriptsize{\color{gray}a} & \scriptsize{\color{gray}b} & \color{green}{\Large{\textbf{c}}} \\
  \color{red}d & \color{blue}e & \scriptsize{\color{gray}f} \\
  \color{blue}g & \color{red}h & \scriptsize{\color{gray}i}
\end{bmatrix}
$$

... and find their determinants.

$$
|A| = \color{green}a \cdot \left| \begin{matrix} \color{red}e & \color{blue}f \\ \color{blue}h & \color{red}i \\ \end{matrix} \right| - \color{green}b \cdot \left| \begin{matrix} \color{red}d & \color{blue}f \\\color{blue}g & \color{red}h \\ \end{matrix}\right| - \color{green}c \cdot \left| \begin{matrix} \color{red}d & \color{blue}e \\ \color{blue}g & \color{red}h \\ \end{matrix}\right|
$$

$$
\\ \LARGE \downarrow
$$

$$
\color{green}a(\color{red}{ei} - \color{blue}{fh}) - \color{green}b(\color{red}{dh} - \color{blue}{fg}) - \color{green}c(\color{red}{dh} - \color{blue}{eg})
$$

### Example

$$
  A = 
\begin{bmatrix}
  6   &   1   &  1  \\
  4   &  \text{-}2   &  5  \\
  2   &   8   &  7  \\
\end{bmatrix}
$$

$$
|A| = \color{green}6 \cdot \left| \begin{matrix} \color{red}{\text{-}2} & \color{blue}5 \\ \color{blue}8 & \color{red}7 \\ \end{matrix} \right|- \color{green}1 \cdot\left|\begin{matrix}\color{red}4 & \color{blue}5 \\\color{blue}2 & \color{red}7 \\ \end{matrix}\right|- \color{green}1 \cdot\left|\begin{matrix}\color{red}4 & \color{blue}{\text{-}2} \\\color{blue}2 & \color{red}8 \\ \end{matrix}\right| \\
$$

$$
\\ \LARGE \downarrow
$$

$$
\color{green}6(\color{red}{\text{-}2}\times \color{red}7 - \color{blue}5\times \color{blue}8) - \color{green}1(\color{red}4\times \color{red}7 - \color{blue}5\times \color{blue}2) - \color{green}1(\color{red}4\times \color{red}8 - \color{blue}{\text{-}2} \times \color{blue}2)
$$

$$
\\ \LARGE \downarrow
$$

$$
\color{green}6(\color{red}{\text{-}14} - \color{blue}{40}) - \color{green}1(\color{red}{28} - \color{blue}{10}) - \color{green}1(\color{red}{32} - \color{blue}{\text{-}4})
$$

$$
\\ \LARGE \downarrow
$$

$$
\color{green}6\times \color{purple}{\text{-}54} - \color{green}1\times \color{purple}{18} - \color{green}1 \times \color{purple}{\text{-}36}
$$

$$
\\ \LARGE \downarrow
$$

$$
\text{-}324 - 18 + 36
\\ \LARGE \downarrow
$$

$$
\text{-}306
$$

## $4 \times 4$ Matrix (and Beyond)

$$
A = \begin{bmatrix} a   &  b   &  c  &  d  \\
                    e   &  f   &  g  &  h  \\
                    i   &  j   &  k  &  l  \\
                    m   &  n   &  o  &  p  \\
    \end{bmatrix}\\
$$

Break it into four $3 \times 3$ matrices...

$$
\small
\begin{bmatrix} 
  \color{darkorange}{\textbf{a}} & \tiny{\color{lightgray}b} & \tiny{\color{lightgray}c} & \tiny{\color{lightgray}d}  \\
  \tiny{\color{lightgray}e}   &  \color{green}f   &  \color{green}g  &  \color{green}h  \\
  \tiny{\color{lightgray}i}   &  j   &  k  &  l  \\
  \tiny{\color{lightgray}m}   &  n   &  o  &  p  \\
\end{bmatrix}
~~
\begin{bmatrix} 
  \tiny{\color{lightgray}a} & \color{darkorange}{\textbf{b}} & \tiny{\color{lightgray}c} & \tiny{\color{lightgray}d}  \\
  \color{green}e   &  \tiny{\color{lightgray}f}   &  \color{green}g  &  \color{green}h  \\
  i   &  \tiny{\color{lightgray}j}  &  k  &  l  \\
  m   &  \tiny{\color{lightgray}n}   &  o  &  p  \\
\end{bmatrix}
~~
\begin{bmatrix} 
  \tiny{\color{lightgray}a} & \tiny{\color{lightgray}b} & \color{darkorange}{\textbf{c}}  & \tiny{\color{lightgray}d}  \\
  \color{green}e   &  \color{green}f  & \tiny{\color{lightgray}g}  &  \color{green}h  \\
  i   &  j  & \tiny{\color{lightgray}k}  &  l  \\
  m   &  n  & \tiny{\color{lightgray}o}  &  p  \\
\end{bmatrix}
~~
\begin{bmatrix} 
  \tiny{\color{lightgray}a} & \tiny{\color{lightgray}b} & \tiny{\color{lightgray}c} & \color{darkorange}{\textbf{d}}  \\
  \color{green}e   &  \color{green}f  &  \color{green}g & \tiny{\color{lightgray}h} \\
  i   &  j  &  k & \tiny{\color{lightgray}l} \\
  m   &  n  &  o  &  \tiny{\color{lightgray}p}
\end{bmatrix}
$$

... then break those into twelve $2 \times 2$ matrices, recursively finding their determinants (minors).

$$
\small
|A| = \color{darkorange}a \cdot \left| \begin{matrix} 
  \color{green}f & \color{green}g & \color{green}h \\
  j & k & l \\
  n & o & p \end{matrix} \right| -
\color{darkorange}b \cdot \left| \begin{matrix} 
  \color{green}e & \color{green}g & \color{green}h \\
  i & k & l \\
  m & o & p \end{matrix} \right| +
\color{darkorange}c \cdot \left| \begin{matrix} 
  \color{green}e & \color{green}f & \color{green}h \\
  i & j & l \\
  m & n & p \end{matrix} \right| -
\color{darkorange}d \cdot \left| \begin{matrix} 
  \color{green}e & \color{green}f & \color{green}g \\
  i & j & k \\
  m & n & o \end{matrix} \right|
$$

$$
\\ \LARGE \downarrow
$$

$$
\scriptsize
\color{darkorange}a \cdot 
\left(
\begin{array}{cc}
\begin{bmatrix}
\color{green}f & \tiny{\color{lightgray}g} & \tiny{\color{lightgray}h} \\
\tiny{\color{lightgray}j} & \color{red}k & \color{blue}l \\
\tiny{\color{lightgray}n} & \color{blue}o & \color{red}p \\
\end{bmatrix} \\ - \\
\begin{bmatrix}
  \tiny{\color{lightgray}f} & \color{green}g  & \tiny{\color{lightgray}h} \\
  \color{red}j & \tiny{\color{lightgray}k} & \color{blue}l \\
  \color{blue}n & \tiny{\color{lightgray}o} & \color{red}p \\
\end{bmatrix} \\ + \\
\begin{bmatrix}
  \tiny{\color{lightgray}f} & \tiny{\color{lightgray}g}  & \color{green}h \\
  \color{red}j & \tiny{\color{lightgray}k} & \color{blue}l \\
  \color{blue}n & \tiny{\color{lightgray}o} & \color{red}p \\
\end{bmatrix}
\end{array}
\right) -
\color{darkorange}b \cdot 
\left(
\begin{array}{cc}
\begin{bmatrix}
\color{green}e & \tiny{\color{lightgray}g} & \tiny{\color{lightgray}h} \\
\tiny{\color{lightgray}i} & \color{red}k & \color{blue}l \\
\tiny{\color{lightgray}m} & \color{blue}o & \color{red}p \\
\end{bmatrix} \\ - \\
\begin{bmatrix}
  \tiny{\color{lightgray}f} & \color{green}g  & \tiny{\color{lightgray}h} \\
  \color{red}i & \tiny{\color{lightgray}k} & \color{blue}l \\
  \color{blue}m & \tiny{\color{lightgray}o} & \color{red}p \\
\end{bmatrix} \\ + \\
\begin{bmatrix}
  \tiny{\color{lightgray}f} & \tiny{\color{lightgray}g}  & \color{green}h \\
  \color{red}i & \color{blue}k & \tiny{\color{lightgray}l} \\
  \color{blue}m & \color{red}o & \tiny{\color{lightgray}p} \\
\end{bmatrix}
\end{array}
\right) +
\color{darkorange}c \cdot 
\left(
\begin{array}{cc}
\begin{bmatrix}
\color{green}e & \tiny{\color{lightgray}f} & \tiny{\color{lightgray}h} \\
\tiny{\color{lightgray}i} & \color{red}j & \color{blue}l \\
\tiny{\color{lightgray}m} & \color{blue}n & \color{red}p \\
\end{bmatrix} \\ - \\
\begin{bmatrix}
  \tiny{\color{lightgray}f} & \color{green}f  & \tiny{\color{lightgray}h} \\
  \color{red}i & \tiny{\color{lightgray}j} & \color{blue}l \\
  \color{blue}m & \tiny{\color{lightgray}n} & \color{red}p \\
\end{bmatrix} \\ + \\
\begin{bmatrix}
  \tiny{\color{lightgray}f} & \tiny{\color{lightgray}f}  & \color{green}h \\
  \color{red}i & \color{blue}j & \tiny{\color{lightgray}l} \\
  \color{blue}m & \color{red}n & \tiny{\color{lightgray}p} \\
\end{bmatrix}
\end{array}
\right) -
\color{darkorange}d \cdot 
\left(
\begin{array}{cc}
\begin{bmatrix}
\color{green}e & \tiny{\color{lightgray}f} & \tiny{\color{lightgray}g} \\
\tiny{\color{lightgray}i} & \color{red}j & \color{blue}k \\
\tiny{\color{lightgray}m} & \color{blue}n & \color{red}o \\
\end{bmatrix} \\ - \\
\begin{bmatrix}
  \tiny{\color{lightgray}e} & \color{green}f  & \tiny{\color{lightgray}g} \\
  \color{red}i & \tiny{\color{lightgray}j} & \color{blue}k \\
  \color{blue}m & \tiny{\color{lightgray}n} & \color{red}o \\
\end{bmatrix} \\ + \\
\begin{bmatrix}
  \tiny{\color{lightgray}e} & \tiny{\color{lightgray}f}  & \color{green}g \\
  \color{red}i & \color{blue}j & \tiny{\color{lightgray}k} \\
  \color{blue}m & \color{red}n & \tiny{\color{lightgray}o} \\
\end{bmatrix}
\end{array}
\right)
$$

$$
\\ \LARGE \downarrow
$$

$$
\scriptsize
\color{darkorange}a \cdot
\left(
\begin{array}{cc}
\color{green}f \cdot
\left| \begin{matrix} 
  \color{red}k & \color{blue}l \\
  \color{blue}o & \color{red}p \\
\end{matrix} \right| \\-\\
\color{green}g \cdot
\left| \begin{matrix} 
  \color{red}j & \color{blue}l \\
  \color{blue}n & \color{red}p \\
\end{matrix} \right| \\+\\
\color{green}h \cdot
\left| \begin{matrix} 
  \color{red}j & \color{blue}k \\
  \color{blue}n & \color{red}o \\
\end{matrix} \right|
\end{array}
\right)
-
\color{darkorange}b \cdot
\left(
\begin{array}{cc}
\color{green}e \cdot
\left| \begin{matrix} 
  \color{red}k & \color{blue}l \\
  \color{blue}o & \color{red}p \\
\end{matrix} \right|
\\-\\
\color{green}g \cdot
\left| \begin{matrix} 
  \color{red}i & \color{blue}l \\
  \color{blue}m & \color{red}p \\
\end{matrix} \right|
\\+\\
\color{green}h \cdot
\left| \begin{matrix} 
  \color{red}i & \color{blue}k \\
  \color{blue}m & \color{red}o \\
\end{matrix} \right|
\end{array}
\right) +
\color{darkorange}c \cdot
\left(
\begin{array}{cc}
\color{green}e \cdot
\left| \begin{matrix} 
  \color{red}j & \color{blue}l \\
  \color{blue}n & \color{red}p \\
\end{matrix} \right| \\-\\
\color{green}f \cdot
\left| \begin{matrix} 
  \color{red}i & \color{blue}l \\
  \color{blue}m & \color{red}p \\
\end{matrix} \right| \\+\\
\color{green}h \cdot
\left| \begin{matrix} 
  \color{red}i & \color{blue}j \\
  \color{blue}m & \color{red}n \\
\end{matrix} \right|
\end{array}
\right) -
\color{darkorange}d \cdot
\left(
\begin{array}{cc}
\color{green}e \cdot
\left| \begin{matrix} 
  \color{red}j & \color{blue}k \\
  \color{blue}n & \color{red}o \\
\end{matrix} \right| \\-\\
\color{green}f \cdot
\left| \begin{matrix} 
  \color{red}i & \color{blue}k \\
  \color{blue}m & \color{red}o \\
\end{matrix} \right| \\+\\
\color{green}g \cdot
\left| \begin{matrix} 
  \color{red}i & \color{blue}j \\
  \color{blue}m & \color{red}n \\
\end{matrix} \right|
\end{array}
\right)
$$

$$
\\ \LARGE \downarrow
$$

$$
\scriptsize
\color{darkorange}a \cdot 
\left(
\begin{array}{cc}
\color{green}f(\color{red}{kp} - \color{blue}{lo})
\\ - \\
\color{green}g(\color{red}{jp} - \color{blue}{ln})
\\ + \\
\color{green}h(\color{red}{jo} - \color{blue}{kn})
\end{array}
\right)
-
\color{darkorange}b \cdot 
\left(
\begin{array}{cc}
\color{green}e(\color{red}{kp} - \color{blue}{lo})
\\ - \\
\color{green}g(\color{red}{ip} - \color{blue}{lm})
\\ + \\
\color{green}h(\color{red}{io} - \color{blue}{km})
\end{array}
\right)
+
\color{darkorange}c \cdot 
\left(
\begin{array}{cc}
\color{green}e(\color{red}{jp} - \color{blue}{ln})
\\ - \\
\color{green}f(\color{red}{ip} - \color{blue}{lm})
\\ + \\
\color{green}h(\color{red}{in} - \color{blue}{jm})
\end{array}
\right)
-
\color{darkorange}d \cdot 
\left(
\begin{array}{cc}
\color{green}e(\color{red}{jo} - \color{blue}{nk})
\\ - \\
\color{green}f(\color{red}{io} - \color{blue}{km})
\\ + \\
\color{green}g(\color{red}{in} - \color{blue}{jm})
\end{array}
\right)
 \\ \LARGE \downarrow
 $$
 
 $$
\scriptsize
\color{darkorange}a(\color{green}f\color{red}{kp} - \color{green}f\color{blue}{lo} - \color{green}g\color{red}{jp} - \color{green}g\color{blue}{ln} + \color{green}h\color{red}{jo} - \color{green}h\color{blue}{kn}) 
-
\scriptsize
\color{darkorange}b(\color{green}e\color{red}{kp} - \color{green}e\color{blue}{lo} - \color{green}g\color{red}{ip} - \color{green}g\color{blue}{lm} + \color{green}h\color{red}{io} - \color{green}h\color{blue}{km}) 
\\\scriptsize+\\
\scriptsize
\color{darkorange}c(\color{green}e\color{red}{jp} - \color{green}e\color{blue}{ln} - \color{green}f\color{red}{ip} - \color{green}f\color{blue}{lm} + \color{green}h\color{red}{in} - \color{green}h\color{blue}{jm}) 
-
\scriptsize
\color{darkorange}d(\color{green}e\color{red}{jo} - \color{green}e\color{blue}{nk} - \color{green}f\color{red}{io} - \color{green}f\color{blue}{km} + \color{green}g\color{red}{in} - \color{green}g\color{blue}{jm})
$$




# R

```{r}
r_mat <- matrix(c(6,  1, 1,  3,
                  4, -2, 5,  9,
                  2,  8, 7,  2,
                  5, -1, 4, -2),
  nrow = 4, ncol = 4, byrow = TRUE
  )
```

## Custom

```{r}
find_determinant <- function(x) {
  stopifnot(is.matrix(x) && dim(x)[[1]] == dim(x)[[2]])
  if (all(dim(x) == 2)) return(x[1, 1] * x[2, 2] - x[1, 2] * x[2, 1])
  
  out <- 0
  
  for (i in seq_len(ncol(x))) {
    out <- out + x[1, i] * (-1)^(1 + i) * find_determinant(x[-1, -i])
  }
  
  out
}

find_determinant(r_mat)
```

## Base R

```{r}
det(r_mat)
```

# Python

## NumPy

```{python}
import numpy as np

np_mat = np.array([[6,  1,  1,  3],
                   [4, -2,  5,  9],
                   [2,  8,  7,  2],
                   [5, -1,  4, -2]])
print(np_mat)
```


```{python}
res = np.linalg.det(np_mat)
print(res)
```


# Rcpp

## RcppArmadillo

```{Rcpp, cache=TRUE}
// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>
using namespace Rcpp;

// [[Rcpp::export]]
double cpp_arma_determinant(const NumericMatrix& x) {
  const arma::mat arma_mat = as<arma::mat>(x);
  
  const double out = arma::det(arma_mat);

  return out;
}
```

```{r}
cpp_arma_determinant(r_mat)
```

<details>
<summary> Sanity Check </summary>
<p>

```{r}
library(reticulate)

stopifnot(
  all.equal(det(r_mat),
            cpp_arma_determinant(r_mat),
            py$np$linalg$det(py$np_mat))
  )
```

</p>
</details>




```{r child=here::here("static/R/environment-details.Rmd")}
```