---
title: EI-Index
draft: true
author: Brendan Knapp
date: '2018-12-03'
slug: index
categories:
  - data-science-notes
tags:
  - data-science-from-scratch
  - network-analysis
  - r
image:
  caption: ''
  focal_point: 'Smart'
summary: "Calculating Krackhardt's EI-Index from scratch."
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
citation_url: https://knapply.com/post/ei-index/
editor_options: 
  chunk_output_type: console
---

> _This post was last updated on `r format(Sys.Date(), "%Y-%m-%d")`._

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, comment = "#>", message = FALSE, warning = FALSE)

options(width=80)

source(here::here("static/R/html-helpers.R"))
source(here::here("static/R/wrangle_plot_save.R"))
```

# Requirements

```{r}
packages_required <- c("ergm", "igraph", "network", "snatools")
packages_available <- installed.packages()[, "Package"]

if (!all(packages_required %in% packages_available)) {
  missings <- packages_required[!packages_required %in% packages_available]
  stop(sprintf("Package(s) Missing:\n%s", paste0(missings, collapse = ", ")))
}
```


# Setup

`igraph` and `network` don't play along, but we can prevent that headache by defining our own methods to wrap the operations we need.

```{r}
# extract vertex attributes
get_attr <- function(x, attr, ...) UseMethod("get_attr")

get_attr.igraph <- function(x, attr, ...) {
  igraph::vertex_attr(graph = x, name = attr, ...)
}

get_attr.network <- function(x, attr, ...) {
  network::get.vertex.attribute(x = x, attrname = attr, ...)
}


# get edgelist representations
get_el <- function(x, ...) UseMethod("get_el")

get_el.igraph <- function(x, ...) {
  igraph::as_edgelist(graph = x, names = FALSE, ...)
}

get_el.network <- function(x, ...) {
  out <- network::as.matrix.network.edgelist(x = x, ...)
  attributes(out) <- attributes(out)[!names(attributes(out)) %in% c("n", "vnames")] 
  out
}
```

## Data

```{r}
data(faux.magnolia.high, package = "ergm")

nw <- faux.magnolia.high
```

```{r}
library(snatools)
ig <- as_igraph(nw)

ig
```


# Vertex Attribute

Using the attribute of interest, get a vector of each vertices' value.

```{r}
attr_of_interest <- "Sex"
```


## `network`

```{r}
nw_attr <- get_attr(nw, attr = attr_of_interest)

head(nw_attr, 15)
```

## `igraph`

```{r}
ig_attr <- get_attr(ig, attr = attr_of_interest)
head(ig_attr, 15)
```

## Single `attr` Variable

```{r}
if (!identical(nw_attr, ig_attr)) {
  stop("attrs don't match")
}

attr <- nw_attr
```


# Index Edge List

Get the edge list as a matrix, filled by vertex indices.

## `network`

```{r}
nw_el <- get_el(nw)

head(nw_el, 10)
```

## `igraph`

```{r}
ig_el <- get_el(ig)

head(ig_el, 10)
```

## Single `el` Variable

```{r}
if (!all.equal(nw_el, ig_el)) {
  stop("edgelists don't match")
}

el <- nw_el
```


# Attribute Edge List

Fill the edge list with the chosen `attr`ibute by using `el` to index into `attr`.

```{r}
vectorized_attr_el <- attr[el]
head(vectorized_attr_el, 15)
```


Give `vectorized_attr_el` `dim`ensions by converting the result to a 2 column `matrix`.

```{r}
attr_el <- matrix(vectorized_attr_el, ncol = 2)
head(attr_el, 10)
```


# Count External and Internal Edges

## External

Filter `attr_el` for `external_rows` by subsetting those where the first column (`attr_el[, 1]`) __does not__ equal  (`!=`) the second column (`attr_el[, 2]`).

```{r}
external_rows <- attr_el[ attr_el[, 1] != attr_el[, 2], ]
head(external_rows, 10)
```

### How many?

```{r}
external_n <- nrow(external_rows)
external_n
```


## Internal

Filter `attr_el` for `internal_rows` by subsetting those where the first column (`attr_el[, 1]`) __does__ equal  (`==`) the second column (`attr_el[, 2]`).

```{r}
internal_rows <- attr_el[ attr_el[, 1] == attr_el[, 2], ]
head(internal_rows, 10)
```

### How many?

```{r}
internal_n <- nrow(internal_rows)
internal_n
```


# Calculate the E-I Index

$$EI~Index = \frac{EL-IL}{EL+IL}$$

```{r}
ei_index <- (external_n - internal_n) / (external_n + internal_n)
ei_index
```


## Validate Result

```{r}
stopifnot(all.equal(ei_index, -0.4147844))
```





```{r child=here::here("static/R/environment-details.Rmd")}
```



