---
title: "Statistical Rethinking: Notes and Practice Question Answers"
draft: true
author: Brendan Knapp
date: '2018-12-05'
slug: index
categories:
  - data-science-notes
tags:
  - statistics
  - bayes
image:
  caption: ''
  focal_point: 'Smart'
summary: "Notes and answers to practice questions from Richard McElreath's _Statistical Rethinking: A Bayesian Course with Examples in R and Stan_."
output:
  # html_document:
  blogdown::html_page:
    toc: true
    toc_depth: 2
citation_url: https://knapply.com/post/statistical-rethinking/
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, comment = "#>", message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
```

# Variables

* $H$: __hypothesis__
* $P$ (or $Pr$): __probability___
* $P(A|B)$: __conditional probability__ of event $A$ given event $B$

# Bayes' Rule

$$P(H|E) = \frac{ P(E|H) \cdot P(H) } { P(E) } = \frac{ P(E|H) } { P(E) } \cdot P(H)$$


* $P(H|E)$: __posterior probability__
    + $P$robability of the $H$ypothesis given the $E$vidence
        + i.e. after $E$ is observed
    + What we want to know: How probable is our hypothesis _given_ the observed evidence?
* $P(H)$: __prior__
    + How probable was our hypothesis _before_ observing the evidence?
    + $P$robability of the $H$ypothesis being true, before gathering $e$vidence
* $P(E|H)$: __likelihood__
    + How probable is observing the $E$vidence, given the $H$ypothesis is true?
* $P(E)$: __marginal likelihood__
    + the (mean) $P$robability of the $E$vidence across all possible hypotheses
    + "model evidence"




# Bayesian Inference

$p$: parameter
$x$: data point or vector of values
$\theta$: parameter of data point's distribution 
    + $x \sim p(x|\theta)$



# Chapter 2: Small Worlds and Large Worlds

__2.3 Components of the model__



$$ Pr(w|n, p) = \frac{n!}{w!(n-2)!}p^w(1-p)^{n-w}$$
$$n_{water~observations} = \frac{all~combinations}{}$$

## Grid Approximation

1. Define the grid
    + How many points to use in estimating posterior?
    + List parameter values on the grid.

```{r}
(p_grid <- seq(from = 0, to = 1, length.out = 20))
```

2. Find prior's value at each parameter value on the grid.

```{r}
(prior <- rep(x = 1, times = 20))
```

3. Find likelihood at each parameter value.

```{r}
(likelihood <- dbinom(x = 6, size = 9, prob = p_grid))
```


4. Find unstandardized posterior at each parameter value...
    + $posterior_{unstandardized} = prior \times likelihood$
    
```{r}
(unstandardized_posterior <- prior * likelihood)
```

5. Standardize posterior by dividing each value by sum of all values
    + $posterior_{standardized} = \frac{posterior_{unstandardized}}{\sum{posterior_{unstandardized}}}$

```{r}
( posterior <- unstandardized_posterior / sum(unstandardized_posterior) )
```


```{r}
plot(p_grid, posterior, type = "b", xlab = "probability of water", ylab = "posterior probability")
```

```{r}
grid_df <- tibble(p_grid = p_grid, 
       likelihood = likelihood,
       unstandardized_posterior = unstandardized_posterior,
       posterior = posterior
       ) 
grid_df %>% 
  knitr::kable(format = "html", escape = FALSE) %>% 
  kableExtra::kable_styling(position = "center") %>% 
  kableExtra::row_spec(0:nrow(grid_df), align = "center")
```


```{r}
plot_grid <- function(...) {
  dots <- list(...)
  df <- data.frame(dots, stringsAsFactors = FALSE)
  col_names <- names(df)
  df %>% 
    ggplot(aes_string(x = col_names[[1]], y = col_names[[2]])) +
    geom_line(color = "dark gray") +
    geom_point() +
    theme_minimal()
}

plot_grid(`probability of water` = p_grid, `posterior probability` = posterior)
```





## Practice

__2E1.__ Which of the expressions below correspond the the statement: _the probability of rain on Monday_?

  1. Pr(rain)
  2. __Pr(rain|Monday)__
  3. Pr(Monday|rain)
  4. Pr(rain, Monday)/Pr(Monday)
    

__2E2.__ Which of the following statements correspond to the expression: Pr(Monday|rain)?
  
  1. The probability of rain on Monday.
  2. The probability of rain, given that is is Monday.
  3. __The probability that it is Monday, given that it is raining.__
  4. The probability that it is Monday and that it is raining.
  
__2E3.__ Which of the expressions below correspond to the statement: _the probability that it is Monday, given that it is raining_?

  1. __Pr(Monday|rain)__
  2. Pr(rain|Monday)
  3. Pr(rain|Monday)Pr(Monday)
  4. Pr(rain|Monday)Pr(Monday)/Pr(rain)
  5. Pr(Monday|rain)Pr(rain)/Pr(Monday)
  
__2E4.__
  
__2M1.__ Recall the globe tossing model from the chapter. Compute and plot the grid approximate posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for $p$.

```{r}
(p_grid <- seq(from = 0, to = 1, length.out = 20))
(prior <- rep(x = 1, times = 20))
(likelihood <- dbinom(x = 6, size = 9, prob = p_grid))
(unstandardized_posterior <- prior * likelihood)
( posterior <- unstandardized_posterior / sum(unstandardized_posterior) )
```





```{r child=here::here("static/R/environment-details.Rmd")}
```
  
  
  