---
title: Variance and Standard Deviation
author: Brendan Knapp
date: '2018-11-23'
slug: index
categories:
  - data-science-notes
tags:
  - data-science-from-scratch
  - statistics
  - programming
  - r
  - python
  - rcpp
image:
  caption: ''
  focal_point: 'Smart'
summary: 'Finding the mode of a sequence, from scratch.'
output:
  blogdown::html_page:
    toc: true
    toc_depth: 5
citation_url: https://knapply.com/post/variance-and-standard-deviation/
editor_options: 
  chunk_output_type: console
---

> _This post was last updated on `r format(Sys.Date(), "%Y-%m-%d")`._

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, comment = "#>", message = FALSE, warning = FALSE)

options(width=90)

source(here::here("static/R/html-helpers.R"))
source(here::here("static/R/citation.R"))
source(here::here("static/R/wrangle_plot_save.R"))
```



# What is it?

## Variance

$$Var(x) = \sigma^2 = \frac{\sum(x_i - \mu_x)^2} {N-1}$$
<center> or... </center>

$$Var(x) = s^2 = \frac{\sum(x_i - \bar{x})^2} {n-1}$$

* $x_i$: for each i in x
* $\mu_x$ and $\bar{x}$: arithmetic mean of the values in $x$
* $N$ and $n$: number of observations in $x$
* $\sum$: sum

### Translation

$$\textbf{variance} ~←~ \frac{\textbf{sum}(~(~\textbf{for}~i~\textbf{in}~x: i -\textbf{mean}(x)~)^2~)} {\textbf{length}(x) - 1}$$

## Standard Deviation

$$\sigma = \sqrt{ \frac{\sum(x_i - \mu_x)^2} {N-1} }$$
<center> or... </center>

$$s = \sqrt{ \frac{\sum(x_i - \bar{x})^2} {n-1} }$$

<center> or, much more simply... </center>

$$\sigma = \sqrt{ Var(x) }$$

### Translation

$$\textbf{standard_deviation} ~←~ \textbf{sqrt}(~\textbf{variance}(x)~)$$

```{r}
library(glue)
```

# R

## Base R

* variance: `r r_as_href("stats::var()", "cor")` 
* standard deviation: `r r_as_href("stats::sd()")`



```{r}
r_vector <- c(1, 4, 3, 2, 4, 5, 4, 3, 4, 6, 4, 3)

glue("variance = {var(r_vector)}
     standard deviation = {sd(r_vector)}")
```



# Rcpp

## Sugar

```{Rcpp}
// [[Rcpp::plugins(cpp17)]]
#include <Rcpp.h>

// [[Rcpp::export]]
double rcpp_sugar_var(const Rcpp::NumericVector& x) {
  return Rcpp::var(x);
}

// [[Rcpp::export]]
double rcpp_sug_sd(const Rcpp::NumericVector& x) {
  return Rcpp::sd(x);
}
```

```{r}
r_vector <- c(4, 1, 4, 3, 2, 5, 4, 3, 4, 6, 4, 3)

glue("variance = {rcpp_sugar_var(r_vector)}
     standard deviation = {rcpp_sug_sd(r_vector)}")
```

## RcppArmadillo

```{Rcpp}
// [[Rcpp::plugins(cpp17)]]
// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>

// [[Rcpp::export]]
auto cpp_arma_var(const Rcpp::NumericVector& x) {
  return arma::var( Rcpp::as<arma::vec>(x) );
}

// [[Rcpp::export]]
auto cpp_arma_sd(const Rcpp::NumericVector& x) {
  return arma::stddev( Rcpp::as<arma::vec>(x) );
}
```

```{r}
r_vector <- c(4, 1, 4, 3, 2, 5, 4, 3, 4, 6, 4, 3)

glue("variance = {cpp_arma_var(r_vector)}
     standard deviation = {cpp_arma_sd(r_vector)}")
```




## Custom Template

```{Rcpp}
// [[Rcpp::plugins(cpp17)]]
#include <Rcpp.h>
#include <cmath> // std::pow

// variance
template <int RTYPE, typename T>                      // variance template
auto cpp_var_(const Rcpp::Vector<RTYPE>& x) {
  const auto n = std::size(x);
  const auto mean = std::accumulate(std::begin(x), std::end(x), 0.0) / n;

  const auto x_minus_x_bar_squared = [mean] (const auto& i) {
    return std::pow(i - mean, 2.0);
  };

  double sum_squared_dev = 0;
  for (auto i : x) {
    sum_squared_dev += x_minus_x_bar_squared(i);
  }

  return sum_squared_dev / (n - 1);
}

// [[Rcpp::export]]
SEXP cpp_var(const SEXP& x) {                         // variance wrapper
  switch( TYPEOF(x) ) {
    case INTSXP:
      return Rcpp::wrap( cpp_var_<INTSXP, int>(x) );
    case REALSXP:
      return Rcpp::wrap( cpp_var_<REALSXP, double>(x) );
  }
  Rcpp::warning("Invalid SEXPTYPE %d (%s).\n", TYPEOF(x), Rcpp::type2name(x));
  return R_NilValue;
}

// standard deviation
template <int RTYPE, typename T>                      // stantard deviation template
auto cpp_sd_(const Rcpp::Vector<RTYPE>& x) {
  return Rcpp::sqrt( cpp_var(x) );
}

// [[Rcpp::export]]
SEXP cpp_sd(const SEXP& x) {                          // standard deviation wrapper
  switch( TYPEOF(x) ) {
    case INTSXP:
      return Rcpp::wrap( cpp_sd_<INTSXP, int>(x) );
    case REALSXP:
      return Rcpp::wrap( cpp_sd_<REALSXP, double>(x) );
  }
  Rcpp::warning("Invalid SEXPTYPE %d (%s).\n", TYPEOF(x), Rcpp::type2name(x));
  return R_NilValue;
}
```


```{r}
r_vector <- c(4, 1, 4, 3, 2, 5, 4, 3, 4, 6, 4, 3)

glue("variance = {cpp_var(r_vector)}
     standard deviation = {cpp_sd(r_vector)}")
```




# Python

## Pure Python

* variance: `r py_as_href("statistics.variance()")`
* standard deviation: `r py_as_href("statistics.stdev()")`

```{python}
import statistics

py_list = [4, 1, 4, 3, 2, 5, 4, 3, 4, 6, 4, 3]

res = {
  "variance": statistics.variance(py_list),
  "standard deviation": statistics.stdev(py_list)
  }

for key, val in res.items():
  print(f"{key}: {val}")
```


## NumPy

* variance: `r py_as_href("np.var()")`
* standard deviation: `r py_as_href("np.std()")`

```{python}
import numpy as np
    
numpy_array = np.array(py_list).astype('float64')

res = {
  "variance": np.var(numpy_array, ddof = 1),
  "standard deviation": np.std(numpy_array, ddof = 1)
  }

for key, val in res.items():
  print(f"{key}: {val}")
```


## CuPy

* variance: `r py_as_href("cp.var()")`
* standard deviation: `r py_as_href("cp.std()")`

```{python}
import cupy as cp
  
cupy_array = cp.array(numpy_array, copy = True)

res = {
  "variance": cp.var(cupy_array, ddof = 1),
  "standard deviation": cp.std(cupy_array, ddof = 1)
  }

for key, val in res.items():
 print(f"{key}: {val}")
```

# Shootout

<details>
<summary> Test Setup </summary>
<p>


```{r, message=FALSE, warning=FALSE}
library(reticulate)
library(tidyverse)
library(bench)
```


Create R Data

```{r}
set.seed(831)
r_data <- seq(-1e4, 1e4, by = 0.2) %>%
  sample(replace = TRUE, size = 1e7)

validation_data <- r_data
```

Transfer to Python

```{python}
from copy import deepcopy

numpy_data = np.array(deepcopy(r.r_data))
cupy_data = cp.array(numpy_data, copy = True)
```


Do all the data match?

```{r validate_data}
```

```{r}
validate_data(
   control_data = validation_data,
  `R Data Used` = r_data,
  `NumPy Data Used` = py$numpy_data,
  `CuPy Data Used` = py$cp$asnumpy(py$cupy_data)
  )
```


```{r}
(n <- scales::comma(length(r_data)))
(mem <- capture.output(pryr::object_size(r_data)))
```

</p>
</details>


<details>
<summary> `bench::mark()` </summary>
<p>

```{r, include=FALSE}
gc()
```


```{r}
res <- mark(
  sd(r_data),
  rcpp_sug_sd(r_data),
  cpp_sd(r_data),
  py_run_string("np.std(numpy_data, ddof = 1)", convert = FALSE),
  py_run_string("cp.std(cupy_data, ddof = 1)", convert = FALSE),
  # min_iterations = 10,
  iterations = 10,
  check = FALSE, filter_gc = FALSE
)
```

</p>
</details>

```{r, echo=FALSE}
as_chunk(here::here("static/R/wrangle_plot_save.R"))
```

<details>
<summary> `wrangle_plot_save()` </summary>
<p>

```{r wrangle_plot_save, eval=FALSE}
```

</p>
</details>

<details>
<summary> `write_gg()` </summary>
<p>

```{r write_gg, eval=FALSE}
```

</p>
</details>

<details>
<summary> `res_gg` </summary>
<p>

```{r, message=FALSE, warning=FALSE}
res_gg <- wrangle_plot_save(res, .n = n, .mem = mem, 
                            .title = "Benchmarks: Standard Deviation")
```



```{r, eval=FALSE}
res_gg
```

</p>
</details>


```{r, echo=FALSE, fig.align='center', fig.width=8, fig.height=10, message=FALSE, warning=FALSE}
res_gg
```


<details>
<summary> Post Test Checks </summary>
<p>

Do all methods obtain the same results?

```{r validate_results}
```


```{r}
validate_results(
   control_res = sd(r_data),
  `Rcpp Sugar` = rcpp_sug_sd(r_data),
   Rcpp =  cpp_sd(r_data),
   NumPy = py$np$std(py$numpy_data, ddof = 1),
   CuPy = py$cp$asnumpy(py$cp$std(py$cupy_data, ddof = 1))
  )
```

Did the input data remain unchanged?

```{r data_still_intact}
```


```{r}
validate_data(
   control_data = validation_data,
  `R Data Used` = r_data,
  `NumPy Data Used` = py$numpy_data,
  `CuPy Data Used` = py$cp$asnumpy(py$cupy_data)
  )
```

</p>
</details>

```{r child=here::here("static/R/environment-details.Rmd")}
```








```{r}
# n_row <- 1e4
# n_col <- 1e4
# 
# set.seed(831)
# r_data <- seq(-1000, 1000, by = 0.1) %>%
#   sample(replace = TRUE, size = n_row * n_col) %>% 
#   matrix(nrow = n_row, ncol = n_col)
```




<!-- ```{python} -->
<!-- import numpy as np -->
<!-- import cupy as cp -->
<!-- from copy import deepcopy -->

<!-- numpy_data = np.array(deepcopy(r.r_data)) -->
<!-- cupy64 = cp.array(numpy_data, copy = True) -->
<!-- cupy32 = cp.array(numpy_data.astype('float32'), copy = True) -->
<!-- ``` -->


<!-- ```{python} -->
<!-- cupy_res32 = cp.std(cupy32, axis=0, ddof = 1) -->
<!-- cupy_res64 = cp.std(cupy64, axis=0, ddof = 1) -->
<!-- ``` -->


<!-- ```{python} -->
<!-- import pycuda.driver as cuda -->
<!-- import pycuda.autoinit -->
<!-- from pycuda.compiler import SourceModule -->

<!-- import skcuda.misc -->

<!-- alloc = pycuda.driver.DeviceAllocation -->

<!-- pycuda_array = pycuda.gpuarray.GPUArray(alloc, dtype = 'float64') -->

<!-- skc.std(  ) -->
<!-- ``` -->



<!-- ```{r} -->
<!-- library(reticulate) -->

<!-- all.equal( -->
<!--   as.vector(py$cp$asnumpy(py$cupy_res32)), -->
<!--   as.vector(py$cp$asnumpy(py$cupy_res64)), -->
<!--   cpp_arma_sd(r_data), -->
<!--   apply(r_data, 2, sd) -->
<!-- ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- res <- bench::mark( -->
<!--   py_run_string("cp.std(cupy32, axis=1, ddof = 1)", convert = FALSE), -->
<!--   py_run_string("cp.std(cupy64, axis=1, ddof = 1)", convert = FALSE), -->
<!--   py_run_string("np.std(numpy_data, axis=1, ddof = 1)", convert = FALSE), -->
<!--   cpp_arma_sd(r_data, by_col = FALSE), -->
<!--   check = FALSE -->
<!--   ) -->
<!-- res -->
<!-- ggplot2::autoplot(res) -->
<!-- ``` -->



